---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 19 @ 11:59PM
author: Jiaheng Cai
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
                      
Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
library(tidyverse)
library(lubridate)
library(miceRanger)
library(caret)
library(randomForest)
library(keras)
library(glmnet)
library(InformationValue)
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in 
the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation 
to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the 
    book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) 
    by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.

MCAR (Missing complete at random): the probability of being missing is the same 
for all cases, we say the data is MCAR. MCAR implies that causes of missing data 
are unrelated to the data. 

MAR (Missing at random): the probability of being missing is the same only within 
groups defined by observed data. 

MNAR (Missing not at random): Not MCAR nor MAR. The probability of being missing 
varies for reasons that are unknown to us.

2. Explain in a couple of sentences how the Multiple Imputation by Chained 
Equations (MICE) work.

First, the algorithm will find a variable B which has strong correlation with 
the variable with missing data A, also another variable C. Regress A on B and 
obtain R^2. Randomly fill the data in the table, the R^2 will change since the 
fill in data are totally random. After that, use random forest to predict A by 
B and C. After, use A and C to predict B. Do this process repeatedly after the 
new R^2 is approaching the original R^2. 

3. Perform a data quality check of the ICU stays data. Discard variables with 
substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by 
`NA`s.

```{r}
icu_cohort_tib = 
  readRDS('/home/odamadafaka/biostat-203b-2021-winter/hw4/icu_cohort.rds') %>% 
  as_tibble() %>%
  subset(select = -c(stay_id.x))

icu_cohort_tib = subset(icu_cohort_tib, select = -c(deathtime, edregtime, 
                                                    edouttime, dod, lactate, arterial_blood_pressure_systolic, arterial_blood_pressure_mean))

#remove false temp
icu_cohort_tib$temperature_fahrenheit[
  icu_cohort_tib$temperature_fahrenheit <= 95] = NA

#remove false res_rate
Q <- quantile(icu_cohort_tib$respiratory_rate, probs = c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$respiratory_rate, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$respiratory_rate[icu_cohort_tib$respiratory_rate <= low |
                                  icu_cohort_tib$respiratory_rate >= up] = NA

#remove false nibpm
Q <- quantile(icu_cohort_tib$non_invasive_blood_pressure_mean, probs = c(.25, .75), 
              na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$non_invasive_blood_pressure_mean, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$non_invasive_blood_pressure_mean[
  icu_cohort_tib$non_invasive_blood_pressure_mean <= low | 
    icu_cohort_tib$non_invasive_blood_pressure_mean >= up] = NA

#remove false nibps
Q <- quantile(icu_cohort_tib$non_invasive_blood_pressure_systolic, probs = 
                c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$non_invasive_blood_pressure_systolic, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$non_invasive_blood_pressure_systolic[
  icu_cohort_tib$non_invasive_blood_pressure_systolic <= low | 
    icu_cohort_tib$non_invasive_blood_pressure_systolic >= up] = NA

#remove false heart rate
Q <- quantile(icu_cohort_tib$heart_rate, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$heart_rate, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$heart_rate[icu_cohort_tib$heart_rate <= low | 
                            icu_cohort_tib$heart_rate >= up] = NA

#remove false wbc
Q <- quantile(icu_cohort_tib$wbc, probs = c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$wbc, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$wbc[icu_cohort_tib$wbc <= low | icu_cohort_tib$wbc >= up] = NA

#remove false hematocrit
Q <- quantile(icu_cohort_tib$hematocrit, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$hematocrit, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$hematocrit[icu_cohort_tib$hematocrit <= low | 
                            icu_cohort_tib$hematocrit >= up] = NA

#remove false hematocrit
Q <- quantile(icu_cohort_tib$hematocrit, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$hematocrit, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$hematocrit[icu_cohort_tib$hematocrit <= low | 
                            icu_cohort_tib$hematocrit >= up] = NA

#remove false sodium
Q <- quantile(icu_cohort_tib$sodium, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$sodium, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$sodium[icu_cohort_tib$sodium <= low | 
                        icu_cohort_tib$sodium >= up] = NA

#remove false potassium
Q <- quantile(icu_cohort_tib$potassium, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$potassium, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$potassium[icu_cohort_tib$potassium <= low | 
                           icu_cohort_tib$potassium >= up]=NA

#remove false magenesium
Q <- quantile(icu_cohort_tib$magnesium, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$magnesium, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$magnesium[icu_cohort_tib$magnesium <= low | 
                           icu_cohort_tib$magnesium >= up]=NA

#remove false glucose
Q <- quantile(icu_cohort_tib$glucose, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$glucose, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5*iqr # Lower Range
icu_cohort_tib$glucose[icu_cohort_tib$glucose <= low | 
                         icu_cohort_tib$glucose >= up]=NA

#remove false cretinine
Q <- quantile(icu_cohort_tib$creatinine, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$creatinine, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$creatinine[icu_cohort_tib$creatinine <= low | 
                            icu_cohort_tib$creatinine >= up]=NA

#remove false chloride
Q <- quantile(icu_cohort_tib$chloride, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$chloride, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5*iqr # Lower Range
icu_cohort_tib$chloride[icu_cohort_tib$chloride <= low | 
                          icu_cohort_tib$chloride >= up]=NA

#remove false calcium
Q <- quantile(icu_cohort_tib$calcium, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$calcium, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$calcium[icu_cohort_tib$calcium <= low | 
                         icu_cohort_tib$calcium >= up] = NA

#remove false bicarbonate
Q <- quantile(icu_cohort_tib$bicarbonate, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(icu_cohort_tib$bicarbonate, na.rm = TRUE)
up <-  Q[2] + 1.5 * iqr # Upper Range  
low <- Q[1] - 1.5 * iqr # Lower Range
icu_cohort_tib$bicarbonate[icu_cohort_tib$bicarbonate <= low | 
                             icu_cohort_tib$bicarbonate >= up] = NA

icu_cohort_tib = subset(icu_cohort_tib, select = -c(creatinine))
summary(icu_cohort_tib)
```

4. Impute missing values by `miceRanger` (request $m=3$ datasets). This step is 
very computational intensive. Make sure to save the imputation results as a file.

```{r}
seqTime <- system.time(
  miceObj <- miceRanger(icu_cohort_tib, m=3, returnModels = TRUE, verbose=FALSE, 
                        max.depth = 10))
dataList <- completeData(miceObj)
head(dataList[[1]])
```

5. Make imputation diagnostic plots and explain what they mean.

```{r}
plotDistributions(miceObj,vars='allNumeric')
```

This density plot compares the original dataset and the imputed dataset. If the 
original numeric data plot is not matching up with the imputed value, it suggests 
that the data is not missing at completely random. 

```{r}
plotCorrelations(miceObj,vars='allNumeric')
```

The boxplot displays the Pearson's correlation coefficient over iteration. It's 
used to determine the convergence over iterations.  

```{r}
plotVarConvergence(miceObj,vars='allNumeric')
```

This plot is used to see the convergence of the imputation. Variables with 
divergent line spread may need more iteration. 
(due to the computational power, I will not do more iterations. )

```{r}
plotModelError(miceObj,vars='allNumeric')
```

This plot also gives the convergence rate of each variable. 

6. Obtain a complete data set by averaging the 3 imputed data sets.
```{r}
data = as.data.frame(data.matrix(dataList$Dataset_1) + 
                       data.matrix(dataList$Dataset_2) + 
                       data.matrix(dataList$Dataset_3)) / 3
# saveRDS(data,str_c('/home/odamadafaka/biostat-203b-2021-winter/hw4/data.rds'))
```

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of 
patients admitted to ICU using demographic information (gender, age, marital 
status, ethnicity), first lab measurements during ICU stay, and first vital 
measurements during ICU stay. For example, you can use (1) logistic regression 
(`glm()` function), (2) logistic regression with lasso penalty (glmnet package), 
(3) random forest (randomForest package), or (4) neural network.

1. Partition data into 80% training set and 20% test set. Stratify partitioning 
according the 30-day mortality status.
```{r}
data = read_rds('data.rds')
train.index <- createDataPartition(data$de30, p = .8, list = FALSE)
train <- data[ train.index,]
test  <- data[-train.index,]
```

2. Train the models using the training set.
```{r}
y_train = as.matrix(train$de30)
x_train = as.matrix(cbind(train$gender, train$anchor_age, train$marital_status,
                          train$ethnicity, train$potassium, train$sodium,
                          train$chloride, train$bicarbonate, train$hematocrit,
                          train$wbc, train$glucose, train$magnesium, train$calcium, 
                        train$heart_rate, train$non_invasive_blood_pressure_mean, 
                        train$non_invasive_blood_pressure_systolic,
                        train$temperature_fahrenheit,train$respiratory_rate))
y_test = as.numeric(test$de30)
x_test = as.matrix(cbind(test$gender, test$anchor_age, test$marital_status,
                          test$ethnicity, test$potassium, test$sodium,
                          test$chloride, test$bicarbonate, test$hematocrit,
                          test$wbc, test$glucose, test$magnesium, test$calcium, 
                        test$heart_rate, test$non_invasive_blood_pressure_mean, 
                        test$non_invasive_blood_pressure_systolic,
                        test$temperature_fahrenheit,test$respiratory_rate))
y_train[y_train == 2] = 0
y_test[y_test == 2] = 0
```

Log:
```{r}
log = glm(as.numeric(y_train) ~ anchor_age + 
                     potassium + sodium + chloride + bicarbonate + hematocrit + 
                     wbc + glucose + magnesium + 
                     calcium+heart_rate 
                      + 
                     non_invasive_blood_pressure_systolic +
                     temperature_fahrenheit+respiratory_rate, data = train, 
          family = "binomial")
summary(log)
```

```{r}
train$de30 = as.factor(train$de30)
rf <- randomForest(de30 ~  gender + anchor_age + marital_status + ethnicity + 
                     potassium + sodium + chloride + bicarbonate + hematocrit + 
                     wbc + glucose + magnesium + 
                     calcium+heart_rate + 
                     non_invasive_blood_pressure_mean + 
                     non_invasive_blood_pressure_systolic +
                     temperature_fahrenheit+respiratory_rate, data = train)
```


3. Compare model prediction performance on the test set.
log:
```{r}
predictionlog = predict(log, test, type="response")
optCutOff = optimalCutoff(y_test, predictionlog)[1] 
predicted.classes = ifelse(predictionlog > optCutOff, "1", "0")
confusionMatrix((as.matrix(predicted.classes)),
                (as.matrix(y_test)))
```
precision: death in 30 days: .082
          live in 30 days: .9919  

Random Forest:
```{r}
rf$confusion[, 'class.error']
predictionrf = predict(rf, test)
confusionMatrix((as.matrix(predictionrf)), (as.matrix(y_test)))
```
precision: death in 30 days: .048
          live in 30 days: .9965  