---
title: "Biostats 203B Homework 1"
author: Jiaheng Cai
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with 
Git and GitHub. Efficient and abundant use of Git, e.g., frequent and 
well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at 
GitHub using your UCLA email.

    **Done**

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`
, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

    **Done**

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two 
branches `master` and `develop`. The `develop` branch will be your main 
playground, the place where you develop solution (code) to homework problems 
and write up report. The `master` branch will be your presentation area. Submit 
your homework files (R markdown file `Rmd`, `html` file converted from R 
markdown, all code and data sets to reproduce results) in `master` branch.

    **Done**

4. After each homework due date, teaching assistant and instructor will check 
out your master branch for grading. Tag each of your homework submissions with 
tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. 
That means if you tag your `hw1` submission after deadline, penalty points will 
be deducted for late submission.

    **Done**

5. After this course, you can make this repository public and use it to 
demonstrate your skill sets on job market.

    **Done**

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data]
(https://mimic-iv.mit.edu), a freely accessible critical care database developed 
by the MIT Lab for Computational Physiology. Follow the instructions at 
<https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI 
`Data or Specimens Only Research` course and (2) obtain the PhysioNet credential 
for using the MIMIC-IV data. Display the verification links to your completion 
report and completion certificate here. (Hint: The CITI training takes a couple 
hours and the PhysioNet credentialing takes a couple days; do not leave it to 
the last minute.)

    [Completion Report]
    (https://www.citiprogram.org/verify/?k1c1a1471-8f58-4652-bcad-0c0e43831872-40412197)

    [Completion Certificate]
    (https://www.citiprogram.org/verify/?w4a2a0203-d3c9-48d0-a40b-0ca6c96e30df-40412197)

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets 
from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of 
data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy 
them into your directory. Do **not** decompress the gz data files. These create 
unnecessary big files on storage and are not big data friendly practices. Just 
read from the data folder `/usr/203b-data/mimic-iv` directly in following 
exercises. 

    **Done**

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the 
functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

    ```{bash}
    ls -al /usr/203b-data/mimic-iv/core
    ls -al /usr/203b-data/mimic-iv/hosp
    ls -al /usr/203b-data/mimic-iv/icu
    ```
    
    Try 
  
    ```{bash,eval=F} 
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz
    ```
  
    After running in my own linux machine, `zcat` command prints out all content 
    inside `admissions.csv.gz` without decompress the file.

    Try
  
    ```{bash,eval=F}  
    zless admissions.csv.gz
    ```
    
    This command is a pager which can be controlled by spacebar and support read 
    backward, also without decompress the file.
  
    Try
    
    ```{bash,eval=F}
    zmore admissions.csv.gz
    ```
  
    This command is a pager which can be controlled by spacebar but does not 
    support read backward, also without decompress the file.
  
    Try
  
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zgrep -c ER admissions.csv.gz
    ```
    
    This command with proper option gives a keyword search result. As this 
    example, how many lines have the word `ER`. This command also perform 
    without unzip file. 
  
4. What's the output of following bash script?
    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
    
    This loop gives detailed information of each `datafile` in the folder `core`
    
    Display the number of lines in each data file using a similar loop.

    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        zcat $datafile | wc -l
      done
    ```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this 
data file? How many unique patients (identified by `subject_id`) are in this 
data file? What are the possible values taken by each of the variable 
`admission_type`, `admission_location`, `insurance`, `language`, 
`marital_status`, and `ethnicity`? Also report the count for each unique value 
of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, 
`uniq`, `wc`, and so on.)

    First few lines and number of rows in `admissions.csv.gz`
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | head
    zcat admissions.csv.gz | wc -l
    ```

    Unique `subject_id`:
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $1}' | tail -n +2 | sort | uniq | wc -l
    ```

    Possible value and count:

    `admission_type`:
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $6}' | tail -n +2 | sort | uniq -c
    ```

    `admission_location`:
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $7}' |tail -n +2 | sort | uniq -c
    ```

    `insurance`:
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $9}' |tail -n +2 | sort | uniq -c
    ```

    `language`
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $10}' |tail -n +2 | sort | uniq -c
    ```

    `marital_status`
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $11}' |tail -n +2 | sort | uniq -c
    ```

    `ethnicity`
    ```{bash}
    cd /usr/203b-data/mimic-iv/core
    zcat admissions.csv.gz | awk -F ',' '{print $12}' |tail -n +2 | sort | uniq -c
    ```
    
## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane 
Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and 
Darcy, your friend thinks that Darcy was the most mentioned. You, however, are 
certain it was Elizabeth. Obtain the full text of the novel from 
<http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local 
folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > 
    pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` 
loop, how would you tabulate the number of times each of the four characters 
is mentioned?

    ```{bash}
    FILE=$(tr -s '[[:punct:][:space:]]' '\n' < pride_and_prejudice.txt)
    elizacount=0
    janecount=0
    lydiacount=0
    darcycount=0
    for w in $FILE
      do
        if [[ $w =~ Jane ]]
        then
                janecount=$((janecount+1))
        elif [[ $w =~ Elizabeth ]]
        then
                elizacount=$((elizacount+1))
        elif [[ $w =~ Lydia ]]
        then
                lydiacount=$((lydiacount+1))
        elif [[ $w =~ Darcy ]]
        then
                darcycount=$((darcycount+1))
        fi
    done
    echo Elizabeth count is $elizacount
    echo Jane count is $janecount
    echo Lydia count is $lydiacount
    echo Darcy count is $darcycount
    ```

2. What's the difference between the following two commands?
    ```{bash}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash}
    echo 'hello, world' >> test2.txt
    ```
    
    First one create a new file and write `hello, world` in it. The second one 
    write `hello, world` and append it into text2.txt (if previously created).

3. Using your favorite text editor (e.g., `vi`), type the following and save the 
file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash}
    cd /home/jihcai/biostat-203b-2021-winter/hw1
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this 
shell script. Why do we need the first line of the shell script?

    The output is the last `"$3"` lines of the first `"$2"` lines of file 
    `"$1"`.
    
    `"$1"`, `"$2"`, and `"$3"` are 3 input values for this script. The command 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ``` 
    passes `pride_and_prejudice.txt` into `"$1"`, 20 into `"$2"` and 5 into 
    `"$3"`.
    
### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, 
`cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, 
`uptime`, `who am i`, `who`, `w`, `id`, `last | head`, 
`echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

    ```{bash}
    cal
    ```

    This one gives the calender of current month.
    
    ```{bash}
    cal 2021
    ```

    This one gives the calender of 2021.

    ```{bash}
    cal 9 1752
    ```
    
    This one gives the calender of September 1752, with missing days.

    ```{bash}
    date
    ```

    This command gives the current UTC time.
    
    ```{bash}
    hostname
    ```

    This one gives the host server name.

    ```{bash}
    arch
    ```

    This one gives the architecture of this server.

    ```{bash}
    uname -a
    ```

    This one gives detailed information of this server.
  
  
    ```{bash}
    uptime
    ```

    This one shows the time since last boot.

    ```{bash}
    who am i
    ```
    
    This command shows the my user information.

    ```{bash}
    who
    ```

    This command shows information of people currently login this server.

    ```{bash}
    w
    ```

    This command shows detailed information of people currently login this 
    server.

    ```{bash}
    id
    ```
    
    This one shows the user id under my name.

    ```{bash}
    last | head
    ```

    This one shows the last 10 users logged or still log in this server.

    ```{bash}
    echo {con,pre}{sent,fer}{s,ed}
    ```

    This one prints out the combination from these 3 bracket.

    ```{bash}
    time sleep 5
    ```

    This command ask the machine to sleep 5 seconds.

    ```{bash}
    history | tail
    ```

    This command shows the last 10 commands inputed in this machine.







