---
title: "Biostat 203B Homework 3"
subtitle: Due Mar 5 @ 11:59PM
author: Jiaheng Cai
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(lubridate)
library(data.table)
```

## Q0. Create MIMIC-IV PostgreSQL database

This section documents how to install PostgreSQL and create a database `mimiciv`. 
You do not need to do anything in this question. The note may be helpful if you 
want do the similar on your own computer. Please do **not** run any code in this 
question on the teaching server, as they are very computation intensive and take 
long.

1. Install PostgreSQL and RPostgreSQL package on CentOS. The postgres in `yum` 
is an old version (v9). We want to install postgres v10 and then build the 
RPostgreSQL package based on it.  
    1. Follow instructions in 
    <https://www.digitalocean.com/community/tutorials/how-to-install-and-use-postgresql-on-centos-7> 
    to install PostgreSQL 10 on CentOS 7.  
    2. Issue following command to install RPostgreSQL in R:
    ```{r, eval = F}
    sudo R -e 'Sys.setenv(PG_INCDIR="/usr/pgsql-10/include/"); 
    Sys.setenv(PG_LIBDIR="/usr/pgsql-10/lib/"); install.packages("RPostgreSQL")'
    ```

2. A PostgreSQL database `mimiciv` was created following the instructions at <https://github.com/MIT-LCP/mimic-iv/tree/master/buildmimic/postgres>. 
    1. Set the psql password for user `postgres` to `postgres`. Read <https://gist.github.com/AtulKsol/4470d377b448e56468baef85af7fd614> regarding 
    postgresql authentication.    
    2. Create database `createdb mimiciv`.  
    3. Create the schemas and tables by `psql -d mimiciv -f create.sql`. This 
    takes a few seconds.  
    3. Import csv.gz files into the `mimiciv` database by 
    `psql -d mimiciv ON_ERROR_STOP=1 -v mimic_data_dir=/usr/203b-data/mimic-iv -f load_gz.sql`. 
    This takes a couple hours.  
    4. Create indexes by `psql -d mimiciv -f index.sql`. 
    This takes less than one hour.  
    5. Set up primary keys `psql -d mimiciv -f constraint.sql`. 
    This takes less than one hour.  
    6. Deposit a subset of labevents data, which contains only ICU stays and the 
    lab items we need, to a table `mimic-labevent-icu` in the `mimiciv` database.
    7. Deposit a subset of chartevents data, which contains only ICU stays and 
    the chart events we need, to a table `mimic-chartevent-icu` in the `mimiciv` 
    database.
    
Steps for 7 & 8:
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
is_teaching_server = Sys.info()[["nodename"]] == "ucla-biostat-203b"
```

Connect to the mimiciv PostgreSQL database.
```{r, eval=FALSE}
# Load configuration settings
# host  <- '127.0.0.1'
# port  <- '5432'
user  <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b", 
                'postgres', # on teaching server
                'odamadafaka')  # on my linux device
password <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b",
                   'postgres', # on teaching server
                   '19981226Cai')  # on my linux device   
dbname <- 'mimiciv'
# Connect to the database using the configuration settings
(con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                  dbname = dbname,
                  user = user, 
                  password = password))
```

Set up paths to mimic data.
```{r,eval=FALSE}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimiciv_path <- "/usr/203b-data/mimic-iv/"
  mimiciv_derived <- "/usr/203b-data/mimic-iv-derived-data/"
} else if (str_detect(os, "Ubuntu")) {
  mimiciv_path <- "/home/odamadafaka/mimic-iv-data/"
  mimiciv_derived <- "/home/odamadafaka/mimic-iv-derived-data/"
}
```

Read in the csv.gz file that contains only items we are interested.
```{r, eval=FALSE}
labevents_filtered_itemid_tib <-
  fread(str_c(mimiciv_derived, "labevents_filtered_itemid.csv.gz"),
        col.names = c("subject_id", 
                      "hadm_id", 
                      "lab_itemid", 
                      "lab_charttime", 
                      "valuenum")) %>%
  as_tibble() %>%
  mutate(lab_charttime = ymd_hms(lab_charttime)) %>%
  print(width = Inf)
```

Read in the ICU stays data.
```{r, eval=FALSE}
icustays_tib <- 
  fread(str_c(mimiciv_path, "icu/icustays.csv.gz")) %>%
  as_tibble() %>%
  mutate(intime = ymd_hms(intime), outtime = ymd_hms(outtime)) %>%
  print(width = Inf)
```

Create a tibble with wanted lab items and their descriptions.
```{r, eval=FALSE}
lab_itemid_desc_tib <- 
  tibble(labitemid = c(50912, 50971, 50983, 50902, 50882, 51221, 
                       51301, 50931, 50960, 50893, 50813),
         labitemdesc = c("creatinine", "potassium", "sodium", "chloride", 
                         "bicarbonate", "hematocrit", "wbc", "glucose", 
                         "magnesium", "calcium", "lactate")) %>%
  print()
```

Only keep the first lab events during ICU stays. Deposit into `mimiciv` database.
```{r, eval=FALSE}
labevents_filtered_itemid_tib %>%
  # only keep subject_id+hadm_id that are in icustays_uniq_tble
  semi_join(icustays_tib, by = c("subject_id", "hadm_id")) %>%
  # pull in the intime of that ICU stay
  left_join(select(icustays_tib, subject_id, hadm_id, intime), 
            by = c("subject_id", "hadm_id")) %>%
  # only keep lab items after ICU intime
  filter(lab_charttime >= intime) %>%
  # group by itemid
  group_by(subject_id, lab_itemid) %>%
  # only keep the first chartime for each item
  # slice_min(lab_charttime, n = 1) %>% # this function takes forever
  # top_n(-1, charttime) %>% # this function takes forever
  arrange(lab_charttime, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  # do not need charttime and intime anymore
  select(-lab_charttime, -intime) %>%
  ungroup() %>%
  # spread
  pivot_wider(names_from = lab_itemid, values_from = valuenum) %>%
  # more informative column names
  rename_at(vars(as.character(lab_itemid_desc_tib$labitemid)), 
            ~lab_itemid_desc_tib$labitemdesc) %>%
  print(width = Inf) %>%
  dbWriteTable(con, "mimic_labevents_icu", ., overwite = TRUE, 
               row.names = FALSE)
```

Read in the shrunk chartevents data.
```{r, eval=FALSE}
chartevents_filtered_itemid_tib <-
  fread(str_c(mimiciv_derived, "chartevents_filtered_itemid.csv.gz"),
        col.names = c("subject_id", 
                      "hadm_id", 
                      "stay_id", 
                      "vitals_charttime", 
                      "vitals_itemid", 
                      "valuenum")) %>%
  as_tibble() %>%
  mutate(vitals_charttime = ymd_hms(vitals_charttime)) %>%
  print(width = Inf)
```

```{r, eval=FALSE}
vitals_itemid_desc_tib <- 
  tibble(vitalitemid = c(220045, 220052, 220181, 220179, 
                         220050, 223761, 220210),
         vitalitemdesc = c("heart_rate", 
                           "arterial_blood_pressure_mean",
                           "non_invasive_blood_pressure_mean", 
                           "non_invasive_blood_pressure_systolic",
                           "arterial_blood_pressure_systolic",
                           "temperature_fahrenheit",
                           "respiratory_rate")) %>%
  print()
```

Deposit into `mimiciv` database.
```{r, eval=FALSE}
chartevents_filtered_itemid_tib %>%
  # only keep subject_id+hadm_id that are in icustays_uniq_tble
  semi_join(icustays_tib, by = c("subject_id", "hadm_id")) %>%
  # pull in the intime of that ICU stay
  left_join(select(icustays_tib, subject_id, hadm_id, intime), 
            by = c("subject_id", "hadm_id")) %>%
  # only keep lab items after ICU intime
  filter(vitals_charttime >= intime) %>%
  # group by itemid
  group_by(subject_id, vitals_itemid) %>%
  # only keep the first chartime for each item
  # slice_min(lab_charttime, n = 1) %>% # this function takes forever
  # top_n(-1, charttime) %>% # this function takes forever
  arrange(vitals_charttime, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  # do not need charttime and intime anymore
  select(-vitals_charttime, -intime) %>%
  ungroup() %>%
  # spread
  pivot_wider(names_from = vitals_itemid, values_from = valuenum) %>%
  # more informative column names
  rename_at(vars(as.character(vitals_itemid_desc_tib$vitalitemid)), 
            ~vitals_itemid_desc_tib$vitalitemdesc) %>%  
  print(width = Inf) %>%
  dbWriteTable(con, "mimic_chartevents_icu", ., overwite = TRUE, 
               row.names = FALSE)
```

Disconnect database and clear workspace.
```{r, eval=FALSE}
dbDisconnect(con)
rm(list = ls())
```

## Q1. Compile the ICU cohort in HW2 Q8 from the PostgreSQL database `mimiciv`. 

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
is_teaching_server = Sys.info()[["nodename"]] == "ucla-biostat-203b"
```

Below is an outline of steps.

1. Connect to database `mimiciv`. We are going to use username `postgres` with 
password `postgres` to access the `mimiciv` database.
```{r}
#Load configuration settings
#host  <- '127.0.0.1'
#port  <- '5432'
user  <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b", 
                'postgres', # on teaching server
                'odamadafaka')  # on my ubuntu device
password <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b",
                   'postgres', # on teaching server
                   '19981226Cai')  # on my ubuntu device   
dbname <- 'mimiciv'
# Connect to the database using the configuration settings
(con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                  dbname = dbname, 
                  user = user, 
                  password = password))
```

2. List all schemas in the `mimiciv` database.
```{r}
dbGetQuery(con, "SELECT SCHEMA_NAME FROM INFORMATION_SCHEMA.SCHEMATA")
```

List tables in the `mimiciv` database:
```{r}
dbListTables(con)
```

List schemas and tables in the `mimiciv` database (bash command).
```{bash}
psql -U odamadafaka -d mimiciv -c "\dt *."
```

3. Connect to the icustays table. Note how to use `Id()` to specify the schema 
containing the table.
```{r}
icustays_tble <- tbl(con, Id(schema = "mimic_icu", table = "icustays")) %>%
  print(width = Inf)
```

4. Connect to the patients table.
```{r}
patients_tble <- tbl(con, Id(schema = "mimic_core", table = "patients")) %>%
  print(width = Inf)
```

5. Connect to the admissions table.
```{r}
admissions_tble <- tbl(con, Id(schema = "mimic_core", table = "admissions")) %>%
  print(width = Inf)
```

6. Connect to the `mimic_labevents_icu` table.
```{r}
labevents_tble <- tbl(con, Id(schema = "public", table = "mimic_labevents_icu")) %>%
  print(width = Inf)
```

7. Connect to `mimic_chartevents_icu` table.
```{r}
chartevents_tble <- tbl(con, Id(schema = "public", table = "mimic_chartevents_icu")) %>%
  print(width = Inf)
```

8. Put things together. Using one chain of pipes `%>%` to perform following data 
wrangling steps: (i) keep only the first ICU stay of each unique patient, (ii) 
merge in admissions and patients tables, (iii) keep adults only (age at 
admission >= 18), (iv) merge in the labevents and chartevents tables, (v) 
display the SQL query, (vi) collect SQL query result into memory as a tibble, 
(vii) create an indicator for 30-day mortality, (viii) save the final tibble to 
an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
icu_cohort = icustays_tble %>% 
  group_by(subject_id) %>%
  filter(rank(intime) == 1) %>%
  ungroup() %>%
  left_join(admissions_tble, by = c('subject_id', 'hadm_id')) %>%
  left_join(patients_tble, by = 'subject_id') %>%
  mutate(age_at_adm = year(admittime) - anchor_year + anchor_age) %>%
  filter(age_at_adm >= 18)%>%
  left_join(labevents_tble, by = c('subject_id', 'hadm_id'))%>%
  left_join(chartevents_tble, by = c('subject_id', 'hadm_id'))%>%
  show_query()%>%
  as_tibble()%>%
  group_by(subject_id) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(de30 = ifelse(.$hospital_expire_flag == 1,'yes','no'))%>%
  saveRDS(str_c('/home/odamadafaka/biostat-203b-2021-winter/hw3/mimiciv_shiny/icu_cohort.rds'))%>%
  print()
```

Close database connection and clear workspace.
```{r}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app 
should reside in the `mimiciv_shiny` folder. The app should provide easy access 
to the graphical and numerical summaries of variables (demographics, 
lab measurements, vitals) in the ICU cohort.